---
title: "Statistics Course Assignment 2 Tooth Growth"
author: "JB"
date: "September 26, 2017"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Overview
The following report describes a basic statistical analysis of the tooth growth data
from the R library.  60 guinea pigs were split into two groups of 30.  One group
was given Vitamin C delivered by orange juice (OJ).  The other group was given
vitamin C delivered as ascorbic acid (VC).  In each group, the guinea pigs were
further divided into three groups given Vitamin C at three different dose rates,
0.5, 1 and 2 mg/day.  This was done to evaluate the effect of Vitamin C dose
rate and delivery method on tooth growth in the guinea pigs.

## Exploratory Analysis
In this section some basic exploratory evaluation of the tooth growth data is done
after it has been loaded.

Install the needed plotting libraries and load the data.  Convert the dose column
to a factor for plotting.
```{r load}
library(ggplot2)
library(gridExtra)
tooth_data <- data.frame(ToothGrowth)
tooth_data$dose <- as.factor(tooth_data$dose)
```

Show the summary of the tooth growth data.
```{r summary}
summary(tooth_data)
```

## Data Analysis
The following section goes into more detail in analyzing the tooth growth data.
Specifically a box and whisker plot is done to show the median, quartiles and range
of the tooth lengths for different delivery methods and dose rates.  Also, hypothesis
testing is done to determine if there are statistically significant differences in tooth
length across different delivery methods and dose rates.  

### Box Plot
The following figure shows a box and whisker plot showing tooth lengths across the three
different dose rates for each of the two Vitamin C delivery methods:  orange juice ("OJ")
and ascorbic acid ("VC").
```{r boxplot, fig.height = 6, fig.width = 12}
box_plot <- ggplot(tooth_data, aes(dose,len)) + geom_boxplot(color="black",
fill="orange")+ facet_grid(.~supp) + 
labs(title="Relationship Between Tooth Length and Vitamin C Dose")+
labs(x=" Vitamin C Dose(mg/day)", y="Tooth length") + theme(text = element_text(size=16))
print(box_plot)
```
**Figure 1:** Box and whisker plot for tooth length for each guinea pig group assigned
a specific delivery method and dose rate.

This figure shows that increasing the Vitamn C dose shows a trend of increasing tooth
length for both delivery methods, but the spread in tooth length overall seems slightly
larger with the ascorbic acid guinea pig group.

### Hypothesis Testing
A two-sample, unpaired Student's t-test will be used to compare tooth length between the
OJ and VC Vitamin C delivery groups. This test was chosen because two small sets of data
will be compared that were collected independently from one another.This test will be 
done both one tailed and two tailed. For the two tailed test, the alternative hypothesis 
is that the mean tooth lengths between the OJ and VC delivery groups are different. For 
the one-tailed test, the alternative hypothesis is that the mean tooth length in the
OJ group is larger than in the VC group.  Only the one-tailed output will be shown since 
the two-tailed test will have a P-value of double the one-tail test result.

Two tailed test:
```{r twotailed_delivery,eval=FALSE}
t.test(tooth_data$len~tooth_data$supp)
```

One-tailed test:
```{r onetailed_delivery}
t.test(tooth_data$len~tooth_data$supp, alternative = "greater")
```

It is also feasible to compare the tooth lengths versus dose rates using this test.
One tailed tests will be used to see if the mean tooth length at higher dose rates
is larger than the tooth length at lower dose rates.

Get the tooth lengths for each dose rate, for both delivery methods.
```{r tooth_dose}
tooth_length_oj_dose1 <-tooth_data$len[tooth_data$dose==1 & tooth_data$supp =="OJ"]
tooth_length_oj_dose0p5 <-tooth_data$len[tooth_data$dose==0.5 & tooth_data$supp =="OJ"]
tooth_length_oj_dose2 <-tooth_data$len[tooth_data$dose==2 & tooth_data$supp =="OJ"]

tooth_length_vc_dose1 <-tooth_data$len[tooth_data$dose==1 & tooth_data$supp =="VC"]
tooth_length_vc_dose0p5 <-tooth_data$len[tooth_data$dose==0.5 & tooth_data$supp =="VC"]
tooth_length_vc_dose2 <-tooth_data$len[tooth_data$dose==2 & tooth_data$supp =="VC"]
```

Do the hypothesis tests comparing the tooth lengths between different dose rates, for
each delivery method.  Only the output of OJ delivery comparing dose levels 1
and 2 mg/day will be shown to save space.
```{r dose_tests not shown, eval=FALSE}
t.test(tooth_length_oj_dose1,tooth_length_oj_dose0p5, alternative = "greater")
t.test(tooth_length_vc_dose1,tooth_length_vc_dose0p5, alternative = "greater")
t.test(tooth_length_vc_dose2,tooth_length_vc_dose1, alternative = "greater")
```

```{r dose_tests shown}
t.test(tooth_length_oj_dose2,tooth_length_oj_dose1, alternative = "greater")

```

## Conclusions
For the delivery comparison, these results show that in the one-tailed case, the p-value
is less than the significance level $\alpha$=0.05, meaning that we can reject the null 
hypothesis and assume that the mean tooth length with OJ is larger than with VC 
delivery. However,the two-tailed test shows that the null hypothesis of there being no
difference in mean tooth lengths between the two delivery methods cannot be rejected 
since the P-value is larger than $\alpha$=0.05.  The two-tailed test is a stricter 
test, and perhaps it shows that more data are required to determine if significant
differences between mean tooth lengths are present for the two Vitamin C delivery 
methods.

For the dose rate comparisons, all P-values showed that increasing dose rate increased
tooth length for a significance level of $\alpha$ = 0.05.  Even if two-tailed tests
had been done, the P-values would be doubled but still less than $\alpha$ = 0.05, so
one could accept the alternative hypothesis that increasing dose rate increases tooth
length, for this significance level.  
