---
title: "Statistics Assignment 1: Central Limit Theorem"
author: "JB"
date: "September 22, 2017"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Overview
This assignment demonstrates the Central Limit Theorem for an exponential
probability density function (pdf).  The idea is that if 40 samples are drawn
1000 times from the exponential distribution and if the mean is taken
across all 40 samples in each draw, then the distribution of the 1000 resulting
sample means should be approximately normally distributed.  The average value
of the sample means should be very close to the theoretical mean of the exponential
pdf.  Also, the standard deviation of the sample mean distribution should
be close to the theoretical standard error, which is the standard deviation of the
exponential pdf divided by the square root of the sample size (n=40 in this
case).  The exponential rate, $\lambda$, will be set to 0.2 for this assignment.
The theoretical mean, $\mu$, and standard deviation, $\sigma$ of the exponential pdf 
is $1/\lambda$.  The sample mean distribution should look completely different (normal,
or bell-shaped) compared to the original exponentional pdf.

##Simulations
This section outlines the code needed to do the analysis.

Load the plotting libraries needed to show the distributions.
```{r loadlibrary}
library(ggplot2)
library(gridExtra)
```

Set $\lambda$, $\mu$ and standard error, $\sigma/\sqrt{n}$.
```{r theoretical parameters}
lambda = 0.2
theoretical_mean <- 1/lambda
standard_error <- (1/lambda)/sqrt(40)
```

For the first plot, we want to show the original theoretical exponential pdf,
but we also want to draw 1000 random variables from the exponential pdf to show
that *without any averaging*, these raw samples are in fact exponentially
distributed.  This code chuck draws those 1000 exponential variables. These
variables will also be put into a data frame for plotting.
```{r exponential}
exp_values <- rexp(1000,lambda)
exp_frame <- data.frame(exp_values)
```

This next code chunk, generates 1000 draws of 40 samples each from the
exponential distribution, and then calculates a sample mean across all 40
samples in each draw.  This results in a vector of 1000 sample means. 
Since an average was taken across all 40 samples in each draw, then 
according to the Central Limit Theorem, this averaging will result in a
vector of sample means that is *normally* (not exponentially) distributed.
The resulting histogram of sample means should therefore have a bell-curve
shape.  The average of all 1000 sample means is calculated, which should agree
closely with the theoretical mean of the exponential distribution determined
previously.  The standard deviation of the sample means should also agree
closely with the expected standard error calculated previously, according
to the Central Limit theorem.  For plotting, the sample means are put into
a data frame.
```{r samplemeans}
mns = NULL
for (i in 1 : 1000){
mns = c(mns, mean(rexp(40, lambda)))
}

average_means <- mean(mns)
sd_means <- sd(mns)
means_frame <- data.frame(mns)
```

##Sample vs. Theoretical Comparison: Mean, Standard Deviation, and Distributions

This section compares the average of the sample mean distribution to the theoretical
mean of the exponential distribution, and they should be very close according to the 
Central Limit Theorem.  The standard deviation of the sample means distribution
should also agree closely to the theoretical standard error, according to this theorem.
The following code summarizes these comparisons:
```{r mean_sd summary}
cat("Theoretical mean: ", theoretical_mean, "\n")
cat("Average of sample means: ", round(average_means,3), "\n")
cat("Theoretical standard error: ", round(standard_error,3), "\n")
cat("Standard deviation of sample means: ", round(sd_means,3), "\n")
```

From these results, the agreement between these quantities is good as expected
from the central limit theorem.  These results are explored more closely in the 
following plot figure.  The plot on the left shows the histogram of the 1000 
original random variable samples from the exponential distribution.  As expected,
they follow this distribution and the theoretical exponential curve is overlaid
on the histogram. $\mu$ is shown as solid blue vertical line on this plot.

In the plot on the right, the density histrogram of the 1000 sample means is shown,
and this is clearly no longer an exponential distribution.  The theoretical normal
distribution with mean, $\mu$ and standard deviation $\sigma/\sqrt{n}$ is overlaid
on the histogram and it fits the sample distribution very well.  The average of the
sample means, $\bar{X}$, is shown as a vertical blue line, and it agrees closely
with $\mu$.  To show that the theoretical standard error and sample standard
deviations agree closely, the theoretical and sample 95% confidence intervals are
shown on the plot.  The theoretical 95% confidence interval, shown by the solid
green lines, is given as $\mu\pm 1.96\sigma/\sqrt{n}$.  The sample 95% confidence
interval, shown as the dashed red lines, is given as $\bar{X}\pm 1.96s$, where $s$
is the sample standard deviation of the sample means vector.  These confidence
intervals agree very closely on the plot.

The following code generates a figure and the two plots side by side.
```{r plots, fig.height = 6, fig.width = 12}
exp_plot <- ggplot(exp_frame, aes(x=exp_values)) +
geom_histogram(alpha = .80, binwidth=1, boundary=1, colour = "black", 
fill = "limegreen",aes(y = ..density..))+
stat_function(fun = dexp, args = list(rate=lambda), size=2)+
geom_vline(xintercept = theoretical_mean, linetype = "solid",color="blue",
size=2,alpha = 0.8)+
labs(title="Sample Exponential Density Plot",x = "x",y="P(x)")+
theme(text = element_text(size=16))

sample_means_plot <- ggplot(means_frame, aes(x=mns)) +
geom_histogram(alpha = .80, bins = 15, colour = "black", 
fill = "orange",aes(y = ..density..))+
stat_function(fun = dnorm, args = list(mean = theoretical_mean, sd = standard_error)
, size=2)+
geom_vline(xintercept = average_means, linetype = "solid",color="steelblue",size=2,
alpha = 0.7)+
geom_vline(xintercept = c(average_means-1.96*sd_means, average_means+1.96*sd_means),
linetype = "dashed",color="red",size=2,alpha=0.6)+
geom_vline(xintercept = c(theoretical_mean-1.96*standard_error, 
theoretical_mean+1.96*standard_error),
linetype = "solid",color="green",size=2,alpha=0.6)+
labs(title="Sample Mean Density Plot",x = "Sample Mean",y="Probability")+
theme(text = element_text(size=16))

plot_CLT <- grid.arrange(exp_plot, sample_means_plot, ncol = 2)
```

**Figure 1:** Density histgram plots of the exponential distribution (left) and 
the sample means normal distribution (right).  The theoretical density functions 
are shown overlaid on the histograms.  The means are shown on both plots.  The
theoretical and sample 95% confidence intervals are also shown on the sample
means plot.

##Conclusion

This exercise emphasizes the Central Limit theorem effectively, demonstrating
that regardless of the shape of the distribution of the original random variables,
averaging samples of these variables over many simulations will produce a data set of
mean values for the variable that is normally distributed.  

